Estimating the Initial Susceptible Fraction in Dengue Dynamics

### 2015-07-25 draft_lm1.tex
LM reviews to the manuscript

Other methods for estimating the  number of susceptibles while 
accommodating time-varying force of infection have been proposed 
before, for measles~\citep{bjornstad_dynamics_2002, 
wallinga_reconstruction_2003}, a disease that shows remarkable seasonality.
These methods try to reconstruct the entire 
series of infectious and susceptibles from case data using deterministic models 
and generally work well for measles because there is a one-to-one 
relationship between exposure and immunity, since measles is caused by a 
single-strain pathogen.
Recently, methods in the same fashion were developed for dengue when 
serotype-specific data is available~\citep{Reiner2014}.
When such data is not available, the series of susceptibles to all possible
serotypes, cannot be reconstructed based solely on a deterministic transmission 
model, since the arrival/re-emergence of  new serotypes, an intrinsically 
stochastic events, can drastically change the pool of susceptibles, throwing off 
any sequential estimation based on the  incidence dynamics.

In this paper, we propose a new approach to estimate the number (fraction) of 
susceptibles using a simplified model of dengue transmission based on a 
single-strain Susceptible-Infectious-Removed (SIR) model with time-varying 
infection rate.
In order to bypass the limitations of not knowing the serotype-specific 
seroprevalence and the exact behaviour of the force of infection through time, 
we
propose to inform the time-varying transmissibility using the \rr series 
derived from the notification data~\citep{nishiura}.
We extend a Bayesian framework previously used to estimate the number 
of susceptibles in Influenza epidemics in Europe~\citep{pone2011} to include 
time-varying force of infection and derive a probability distribution  for 
\rr to accommodate uncertainty in the estimates. 
Then, from the incidence series and the population at risk, we  calculate the 
attack ratio for each epidemic.
We apply our method to estimate $S_0$ before every major dengue epidemic in the 
city of Rio de Janeiro, Brazil in the last 18 years. Dengue has been 
reintroduced in Rio de Janeiro in 1986 after being absent for 68 
years\citep{nogueira1999dengue}. During the last decades of the 20th century 
only DEN-1 and DEN-2 serotypes

@@

We explore the approach of Ederer and Mantel~\citep{mantel}, whose objective is 
to obtain 
confidence intervals for the ratio of two Poisson counts. 
Let $Y_{t} \sim Poisson(\lambda_t)$ and $Y_{t+1} \sim Poisson(\lambda_{t+1})$ 
and define $W = Y_{t} + Y_{t+1}$.
The authors note that by conditioning on the sum $W$
\begin{align}
\label{eq:binlike}
Y_{t+1} | S &\sim Binomial(W, \theta_t) \\
\theta_t &= \frac{\lambda_{t+1}}{\lambda_{t} + \lambda_{t+1}}
\end{align}
Let $c_{\alpha}(\theta_t) = \{\theta_t^{(L)} , \theta_t^{(U)} \}$ be such that 
$Pr(\theta_t^{(L)}<\theta_t <\theta_t^{(U)}) = \alpha$.
Analogously, define $c_{\alpha}(R_t) = \{R_t^{(L)} , R_t^{(U)} \}$ such that 
$Pr(R_t^{(L)}<R_t<R_t^{(U)}) = \alpha$.
Ederer and Mantel (1974)~\citep{mantel} show that one can construct a 
$100\alpha 
\%$ confidence interval for \rr by noting that
\begin{equation}
\label{eq:confRt}
 R_t^{(L)} = \frac{\theta_t^{(L)}}{(1-\theta_t^{(L)})} \quad \text{and} \quad 
R_t^{(U)} = \frac{\theta_t^{(U)}}{(1-\theta_t^{(U)})}\\
\end{equation}
Because the transform from $\theta$ to \rr is monotonically 
increasing, the result holds for confidence and credibility intervals alike.

@@

In this paper we propose a method to bypass the lack of serotype-specific case 
data by informing the time-varying force of infection with the instantaneous 
reproductive number, \rr which we calculate from aggregated data.
The main contribution of this paper can be summarized in the following items: 
(i) we show a method to quantify uncertainty 
about \rr that is readily applicable to other diseases and; (ii) we propose to 
use \rr to inform a dynamic epidemic model with time-varying force of infection 
in order to gain insight into the attack ratio of each epidemic; (iii) We 
propose an estimation procedure for circulating serotype's $S_0$ from 
aggregate case data, which is robust to epidemic sizes; (iv) We estimate the AR 
for 18 years of Dengue epidemics.

@@

A key epidemiological quantity is the attack ratio (AR) of an epidemic, a 
measure of morbidity and speed of spread which can be used to predict epidemic 
size and help efficient Public Health planning.
The AR depends fundamentally on the population at risk, which in the case of 
dengue is every naive (to a particular serotype) individual in the population.
Estimating the initial susceptible fraction $S_0$ for each epidemic is thus 
central to the estimation of the AR.
Methods for estimating the  number of susceptibles have been proposed 
before, for other diseases~\citep{bjornstad_dynamics_2002, 
wallinga_reconstruction_2003}.
These methods try to reconstruct the entire 
series of infectious and susceptibles for measles 
outbreaks from case data.
In the case of dengue, the full (multi-year/multi-epidemic)
series of susceptibles to all possible serotypes, cannot be reconstructed based 
solely on a deterministic transmission model, since the arrival/re-emergence of 
new serotypes (which are a stochastic events) can change drastically the pool 
of susceptibles throwing off any sequential estimation based on the incidence
dynamics.

@@

Table~\ref{tab:AR} contains the attack ratios and medians of the $S_0$ 
estimated for each epidemic/outbreak.
 Underreporting of cases, which is known to exist but of which exact figures 
cannot be determined, will lead to underestimation $S_0$. However, The attack 
ratio shall remain unbiased as the underreporting affects both the numerator and 
the denominator of equation~\ref{eq:AR2}.
It is interesting to notice that the larger epidemics, in terms of peak size 
are not the one with the greater attack ratios.
This stresses the importance of knowing the immunological structure of the 
population. Knowing the $S_0$ for the circulating viruses we can order to more 
accurately assess the potential impact of a coming epidemic, since particularly 
virulent types, can be rendered less of a threat by a low $S_0$. Hon\'orio et 
al.\citep{honorio2009spatial} conducted a serological survey in three separate 
localities within the city, right before the 2008 epidemic, the authors report 
seroprevalences varying from 56-77.4\% which is compatible with our prediction 
of 87.5\% ($1-S_{0,2007}$) for the entire city, considering that we 
underestimate $S_0$ due to the underreporting of cases.

@@

>>>>

Other methods for estimating the  number of susceptibles while 
accommodating time-varying force of infection have been proposed 
before, for measles~\citep{bjornstad_dynamics_2002, 
wallinga_reconstruction_2003}, a disease that shows remarkable seasonality.
These methods try to reconstruct the entire 
series of infectious and susceptibles from case data using deterministic models 
and generally work well for measles because there is a one-to-one 
relationship between exposure and immunity, since measles is caused by a 
single-strain pathogen.
Recently, methods in the same fashion were developed for dengue when 
serotype-specific data is available~\citep{Reiner2014}.
When such data is not available, the series of susceptibles to all possible
serotypes, cannot be reconstructed based solely on a deterministic transmission 
model, since the arrival/re-emergence of  new serotypes, an intrinsically 
stochastic event, can drastically change the pool of susceptibles, throwing off 
any sequential estimation based on the  incidence dynamics.

Dengue has been reintroduced in Rio de Janeiro in 1986 after being absent for 
68 years~\citep{nogueira1999dengue}.
During the last decades of the 20th century only DEN-1 and DEN-2 serotypes

In this paper, we propose a new approach to estimate the number (fraction) of 
susceptibles using a simplified model of dengue transmission based on a 
single-strain Susceptible-Infectious-Removed (SIR) model with time-varying 
infection rate.
In order to bypass the limitations of not knowing the serotype-specific 
seroprevalence and the exact behaviour of the force of infection through time, 
we propose to inform the time-varying transmissibility using the \rr series 
derived from the notification data~\citep{nishiura}.
We extend a Bayesian framework previously used to estimate the number 
of susceptibles in Influenza epidemics in Europe~\citep{pone2011} to include 
time-varying force of infection and derive a probability distribution  for 
\rr to accommodate uncertainty in the estimates. 
Then, from the incidence series and the population at risk, we  calculate the 
attack ratio for each epidemic.
We apply our method to estimate $S_0$ before every major dengue epidemic in the 
city of Rio de Janeiro, Brazil in the last 18 years.

@@

We explore the approach of Ederer and Mantel~\citep{mantel}, whose objective 
was  to obtain confidence intervals for the ratio of two Poisson counts. 
Let $Y_{t} \sim Poisson(\lambda_t)$ and $Y_{t+1} \sim Poisson(\lambda_{t+1})$ 
and define $W = Y_{t} + Y_{t+1}$.
The authors note that by conditioning on the sum $W$
\begin{align}
\label{eq:binlike}
Y_{t+1} | S &\sim Binomial(W, \theta_t) \\
\theta_t &= \frac{\lambda_{t+1}}{\lambda_{t} + \lambda_{t+1}}
\end{align}
Let $c_{\alpha}(\theta_t) = \{\theta_t^{(L)} , \theta_t^{(U)} \}$ be such that 
$Pr(\theta_t^{(L)}<\theta_t <\theta_t^{(U)}) = \alpha$.
Analogously, define $c_{\alpha}(R_t) = \{R_t^{(L)} , R_t^{(U)} \}$ such that 
$Pr(R_t^{(L)}<R_t<R_t^{(U)}) = \alpha$.
Ederer and Mantel (1974)~\citep{mantel} show that one can construct a 
$100\alpha 
\%$ confidence interval for \rr by noting that
\begin{equation}
\label{eq:confRt}
 R_t^{(L)} = \frac{\theta_t^{(L)}}{(1-\theta_t^{(L)})} \quad \text{and} \quad 
R_t^{(U)} = \frac{\theta_t^{(U)}}{(1-\theta_t^{(U)})}\\
\end{equation}
Because the transform from $\theta$ to \rr is monotonically 
increasing, the result holds for confidence and credibility intervals alike.

@@

In this paper we propose a method to bypass the lack of serotype-specific case 
data by informing the time-varying force of infection with the instantaneous 
reproductive number, \rr which we calculate from aggregated data.
The main contribution of this paper can be summarized in the following items: 
(i) we develop a method to quantify uncertainty 
about \rr that is readily applicable to other diseases and; (ii) \rr is used to 
inform a dynamic epidemic model with time-varying force of infection 
in order to gain insight into the attack ratio of each epidemic; (iii) we 
propose an estimation procedure for circulating serotype's $S_0$ from 
aggregate case data, which is robust to epidemic sizes; (iv) AR estimates are 
provided for 18 years of Dengue epidemics in Rio de Janeiro, Brazil.

@@

A key epidemiological quantity is the attack ratio (AR) of an epidemic, a 
measure of morbidity and speed of spread which can be used to predict epidemic 
size and help efficient Public Health planning.
The AR depends fundamentally on the population at risk, which in the case of 
dengue is every naive (to a particular serotype) individual in the population.
Estimating the initial susceptible fraction $S_0$ for each epidemic is thus 
central to the estimation of the AR.
Methods for estimating the  number of susceptibles have been proposed 
before, for other diseases~\citep{bjornstad_dynamics_2002, 
wallinga_reconstruction_2003}.
These methods attempt to reconstruct the entire 
series of infectious and susceptibles for measles 
outbreaks from case data.
In the case of dengue, the full (multi-year/multi-epidemic)
series of susceptibles to all possible serotypes, cannot be reconstructed based 
solely on a deterministic transmission model, since the arrival/re-emergence of 
new serotypes (which are a stochastic events) can change drastically the pool 
of susceptibles throwing off any sequential estimation based on the incidence
dynamics.

@@

Table~\ref{tab:AR} contains the attack ratios and medians of the $S_0$ 
estimated for each epidemic/outbreak.
 Underreporting of cases, which is known to exist but of which exact figures 
cannot be determined, will lead to underestimation $S_0$.
However, the attack ratio shall remain unbiased as the underreporting affects 
both the numerator and the denominator of equation~\ref{eq:AR2}.
It is interesting to notice that the larger epidemics, in terms of peak size 
are not the one with the greater attack ratios.
This stresses the importance of knowing the immunological structure of the 
population.
Knowing the $S_0$ for the circulating viruses we can order to more 
accurately assess the potential impact of a coming epidemic, since particularly 
virulent types, can be rendered less of a threat by a low $S_0$.
Hon\'orio et al.\citep{honorio2009spatial} conducted a serological survey in 
three separate localities within the city, right before the 2008 epidemic, the 
authors report seroprevalences varying from 56-77.4\% which is compatible with 
our prediction of 87.5\% ($1-S_{0,2007}$) for the entire city, considering that 
we underestimate $S_0$ due to the underreporting of cases.

@@

\section*{Estimating $S_0$ from simulated data}

In order to determine whether the inference methodology proposed can 
recover the true parameter values of the underlying simulation model, we have 
devised a simple simulation experiment.
Using the SIR model presented in the main paper we have simulated the incidence 
curve of the  2013 epidemic (2012 in table~\ref{eq:AR2}), using 
$S_0 = 0.0621$ and \rr, as estimated from actual incidence data, and $\tau=1$. 
Figure~\ref{fig:sim_data} shows the posterior $S$ and $I$ alongside with 
simulated incidence data.
It is clear that the method can recover the correct value for $S_0$ 
used to   simulate incidence.
The script to generate the simulated data and run the inference is available at 
the paper's Github repository 
(\url{
https://github.com/fccoelho/paperLM1/blob/master/python/fit_simulated_data.py}).

\begin{figure}
 \centering
 \includegraphics[width=14cm]{./plots/Sim_DengueS2012_11_series.png}
 % Sim_DengueS2012_11_series.png: 0x0 pixel, 300dpi, 0.00x0.00 cm, bb=
 \caption{$S$ and $I$ series fitted to simulated incidence data (blue dots) 
with $S_0=0.0621$, and \rr estimated from real incidence data.}
 \label{fig:sim_data}
\end{figure}
